An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-30 DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
finished description of PISA part

In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure during initialization and variation that all neighbored vertices in the path are connected by an arc (see Sec.~\ref{sec:init} and \ref{sec:variation}), and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$. The right hand side of Fig.~\ref{fig:paths} shows the entire source--destination paths corresponding to the representation in the left part of the figure.

@@

\subsection{Initialization}\label{sec:init}
Often, the population of a (multiobjective) evolutionary algorithm is initialized with solutions that are drawn uniformly at random within the search space. Although it is not easy to define a uniformity measure in our case of variable-length representation, we still propose a random initialization scheme for our evolutionary algorithm. In addition, the specific problem formulation allows another initialization where a Pareto-optimal solution is inserted in the initial population. In the following, we present both approaches in more detail and see later on how they compare with respect to the overall performance of the algorithm.

\subsubsection{A Simple Random Initialization}
Based on the idea of representing a solution by the beginning of the truck paths which are completed by the shortest path to the destination nodes, we propose the following random initialization:
\begin{enumerate}
	\item Choose for each truck a node in the network uniformly at random among all nodes which are not the source.
	\item Initialize the truck paths with the shortest path from the associated source node to the randomly selected node.
\end{enumerate}
Note that this initialization scheme is, though random, does not produce fully random paths. The initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.

@@

\subsection{Variation}\label{sec:variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}

$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.

 The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to

\TODO{describe how PISA works}

>>>>

In order to have every variable length path represent an uninterrupted path from source to destination at any time (see the constraints in (\ref{eq:flow})), we ensure all paths to always start with the source $s^c$ for the corresponding commodity $c$, ensure during initialization and variation that all neighbored vertices in the path are connected by an arc (see Sec.~\ref{sec:init} and \ref{sec:variation}), and complete each path by the shortest path between the path's actual end node and the commodity's destination node $t^c$. The shortest path is thereby calculated exactly with respect to the given costs of the arcs. The right hand side of Fig.~\ref{fig:paths} shows the entire source--destination paths corresponding to the representation in the left part of the figure.

@@

\subsection{Initialization}\label{sec:init}
Often, the population of a (multiobjective) evolutionary algorithm is initialized with solutions that are drawn uniformly at random within the search space \citep{deb2001a}. Although it is not easy to define a uniformity measure in our case of a variable-length representation, we still propose a random initialization scheme for our evolutionary algorithm. In addition, the specific problem formulation allows another initialization where a Pareto-optimal solution is inserted in the initial population. In the following, we present both approaches in more detail and see later on how they compare with respect to the overall performance of the algorithm.

\subsubsection{A Simple Random Initialization}
Based on the idea of representing a solution by the beginning of the truck paths which are completed by the shortest path to the destination nodes, we propose the following random initialization:
\begin{enumerate}
	\item Choose for each truck a node in the network uniformly at random among all nodes which are not the source.
	\item Initialize the truck paths with the shortest path from the associated source node to the randomly selected node.
\end{enumerate}
Note that this initialization scheme, though random, does not produce fully random paths. Instead, the initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.

@@

\subsection{Variation}\label{sec:variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as suggested and analyzed in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors provided by PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{abbz2009b,bbtz2011a} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals by the variator, their evaluation and sending the objective values to the selector module, a standard PISA run always continues in the same way. The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector. In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted. In the case of the selectors used in this study, no archive is maintained. The search stops after a predefined number of variator/selector iterations, also called generations, are performed. In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter. For more details about PISA, we refer to \citep{bltz2003a}.

\subsection{The HypE and NSGA-II Selectors of PISA}
\TODO{write}
